Probabilistic Programming

### 2014-04-21 paper/paper.tex

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}

@@

\usepackage[]{algorithm2e}

@@

@@

Other applications of the model: genetic algorithms

@@

The vim autocomplete script

The guided IDE

Design decisions, HCI factors

@@

\subsection{Limitations of the IDE}

Usability issues

@@

Choosing valid identifiers, matching identifiers

@@

\clearpage
\begin{thebibliography}{1}

	\bibitem{esprima} Ariya Hidayat {\em Esprima}
		URL = \url{http://esprima.org/}

	\bibitem{markov} Takis Konstantopoulos {\em Introductory lecture notes on
		Markov Chains and Random Walks} Uppsala University,
		URL = \url{http://www2.math.uu.se/~takis/L/McRw/mcrw.pdf}

	\bibitem{escodegen} Yusuke Suzuki {\em Escodegen}
		URL = \url{https://github.com/Constellation/escodegen}

	\bibitem{parser api} Mozilla Developer Network {\em SpiderMonkey Parser API}
		URL = \url{https://developer.mozilla.org/en-US/docs/SpiderMonkey/Parser_API}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}

@@

\usepackage[]{algorithm2e}
\hyphenation{ECMAScript JavaScript escodegen esprima}

@@

See figure~\ref{fig:sample-model} in the appendix for a model for the program
``var x = 34;" In this model we see for example that the path
\texttt{[Program, body, VariableDeclaration, declarations]} leads to
\texttt{VariableDeclarator} with probability 1. This indicates that a Program
node whose body is a VariableDeclaration always has a declaration that is of
type VariableDeclarator.

The model is implemented as a JSON object where the path is indexed into the
object by successively indexing by each string of the path.

@@

@@

\subsection{vim autocomplete script}

We integrated our model into the text editor vim to offer autocompletion
suggestions of programs upon the user's request, using vim's
\texttt{completefunc} feature. When the user is editing a JS file in vim with
this feature, they may press a keyboard command to get a list of program
snippets to choose from. The snippets are generated in the context of the
program path at cursor's position. The user may pick one of the snippet
suggestions to have it inserted into their code at their cursor.

The autocomplete snippets are generated as follows:
\begin{enumerate}
	\item Parse the contents of the current file into an AST.
	\item Find the node path corresponding to the cursor by traversing the AST.
	\item Generate programs starting from the node path of the cursor.
	\item Present the programs to the user as code completion snippets.
\end{enumerate}

\begin{figure}[ht]
	\centering
	\begin{tabular}{c c}
		\begin{subfigure}[h]{6.25cm}
			\centering
			\includegraphics[width=1.00\textwidth]{vim1}
			\caption{Inserting a for loop} \label{fig:vim1}
		\end{subfigure}
		\begin{subfigure}[h]{6cm}
			\centering
			\includegraphics[width=1.00\textwidth]{vim2}
			\caption{Inserting an object property} \label{fig:vim2}
		\end{subfigure}
	\end{tabular}
	\caption{Screenshots of vim using autocomplete of programs}
\end{figure}

The vim autocomplete feature is limited to inserting program snippets on a single
line, and requires the user to explicitly trigger the \texttt{completefunc}
feature to be prompted with code snippets. To further explore using our model
for editing, we created a web-based editor that more deeply integrates Probabilistic
Programming.

\subsection{Guided IDE}

The Guided IDE offers users an environment for building a program entirely using
our model, using program snippets generated at the cursor's context as building
blocks. The editor presents a window for the program text, and and list of
suggestion code snippets. The user can move a cursor through the code window at
positions corresponding to boundaries of program nodes. As the user moves the
cursor, the snippet suggestions are regenerated using the cursor's node path, to
continuosly provide suggestions relevant to the current context of the program. When the
user selects a suggestion snippet, its AST is inserted into the program's AST at
the cursor's position in the tree, and the code window is redrawn. The user can
move the cursor within newly inserted blocks, so that the program can grow
deeper. The initial corpus for the IDE's model consists of the JavaScript files
that comprise our project.

\begin{figure}[h]
  \centering
  \includegraphics[width=1.00\textwidth]{screenshot}
  \caption{Screenshot of the web program editor GUI} \label{fig:screenshot}
\end{figure}

@@

We can use the Guided IDE to generate programs. The programs look like actual
programs but do not actually function unless the user is very careful about
which snippets they pick. Crucially, the variable and identifier names that our
system generates are not in general meaningful in the scope of the generated
program.  Whereas with the vim autocompletions the identifiers would often be
valid because because the corpus was the current file, with the larger corpus of
the Guided IDE and writing a program from scratch, the user is less likely to
find program snippets that use valid identifiers.

@@

It is a limitation of the current model that identifier names are chosen
probabilistically based on their position in the code, without regard to whether
they are in scope and without refering to anything in another part of the
program. Because of this limitation, the Guided IDE is not useful for practical
purposes. We consider some alternative approaches for the Guided IDE using a
similar model.

\subsection{Alternative approaches for the IDE}

\subsubsection{Prompt for identifiers}

Rather than picking identifier names from the model, use placeholder identifiers
and prompt the user to enter actual identifiers, with suggestions based on the
current variables in scope and previously used identifiers. This would give the
user the opportunity to create semantically valid programs, but the entering of
identifier names could be tedious.

\subsubsection{Stubs}

Instead of having a movable cursor in the code window, have stubs in the
program. Each stub would be a marker in place of an AST node. The user would
select a stub and be given choices of snippets to replace it. Selecting a
snippet would move the current snippet into the program in place of the selected
stub.  The generated snippets could themselves have stubs in them, so that the
user may continue to complete stubs and expand their program.  At the level of
variable identifiers, the snippet suggestions would be taken from the computed
list of in-scope variables at the position of the selected stub, unless the
completion was for a variable declaration, in which case the user would be
prompted to name a new variable.

This approach, while offering less places to insert code then the current
approach, would allow for programs to more closely be generated from our model,
since the stub completions would always be at leaf nodes of the AST, whereas in
the current approach the user may choose to insert a snippet in the middle of a
list, which means that the transition from the inserted snippet to the following
node in the list does not in general follow from the model.

While the stub approach would be more restrictive for the user as it requires
them to build the program in a certain way, with proper handling of identifiers,
it may provide for an interesting or useful paradigm for interacting with a program
source.

@@

%Other applications of the model: genetic algorithms

\subsection{Extending the model}

Just as our approach of modeling a program with a tree-based Markov chain is a
higher order model than a linear Markov chain of the program source code tokens,
there may be applications of probabilistic modeling to higher-order
representations of programs than the AST, such as a higher-order abstract syntax
\cite{hosa}. Such an application may model the relationship between different
parts of an AST, and handle scoping of variables to model identifiers
abstractly.

With a more sophisticated model of identifiers that could generate programs with
identifiers in meaningful relation to eachother, Probabilistic Programming could
be used to generate semantically valid programs, for greater assistive,
expressive, and programmatic power.

\begin{thebibliography}{1}

	\bibitem{esprima} Ariya Hidayat {\em Esprima}
		\\\url{http://esprima.org/}

	\bibitem{markov} Takis Konstantopoulos {\em Introductory lecture notes on
		Markov Chains and Random Walks.} Uppsala University,
		\\\url{http://www2.math.uu.se/~takis/L/McRw/mcrw.pdf}

	\bibitem{hosa} Frank Pfenning, Conal Elliott
		{\em Higher-Order Abstract Syntax.}
		ACM SIGPLAN Notices. Vol. 23. No. 7. ACM, 1988.
		\\\url{http://www.cs.cmu.edu/~fp/papers/pldi88.pdf}

	\bibitem{escodegen} Yusuke Suzuki {\em Escodegen}
		\\\url{https://github.com/Constellation/escodegen}

	\bibitem{parser api} Mozilla Developer Network {\em SpiderMonkey Parser API}
		\\\url{https://developer.mozilla.org/en-US/docs/SpiderMonkey/Parser_API}
